---
title: "LLM Review With Context Files"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LLM Review With Context Files}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE, purl = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

Use this guide when you want `create_sdp()` or `suggest_semantics()` to review semantic candidates with an LLM **and** you have supporting files such as README notes, data dictionaries, or technical reports.

## What Context Files Are Supported

`llm_context_files` accepts local files that can add domain context to the LLM review step:

- markdown/text notes: `.md`, `.txt`, `.rst`
- delimited/text data: `.csv`, `.tsv`, `.json`, `.yaml`, `.yml`
- source/notebook files: `.R`, `.Rmd`, `.qmd`
- HTML pages: `.htm`, `.html`
- Word documents: `.docx`
- PDF reports: `.pdf` with the optional `pdftools` package
- Excel workbooks: `.xls`, `.xlsx`, `.xlsm` with the optional `readxl` package

The files are read locally, chunked, and trimmed before prompting. They are used as supporting evidence only. The LLM still has to choose from the deterministic shortlist returned by `find_terms()`; it does not mint raw IRIs.

## Input Contract and Explicit Opt-in

`llm_context_files` accepts a character vector of existing local **file paths**.
Keep the paths themselves instead of reading the files first:

```{r, purl = FALSE}
context_files <- c(
  "./00_data/Data_dictionary_final_dataset.rmd",
  "./00_data/Data_dictionary_final_dataset.html",
  "./00_data/Data_dictionary_final_dataset.csv"
)
```

Do not pass the result of `readr::read_csv()`, `xml2::read_html()`, an R Markdown
reader, or another parsed object. A tibble or XML document is data in memory, not
a path, and now fails early with a path-specific error. Use `llm_context_text`
when the context is already available as inline character text.

Context is also strictly opt-in: `llm_context_files` and `llm_context_text` never
trigger a network or LLM call. Set `llm_assess = TRUE` explicitly. If context is
supplied without it, `metasalmon` warns that the context is ignored and continues
with deterministic retrieval, so similar output is expected.

When calling `infer_dictionary()` directly, both `seed_semantics = TRUE` and
`llm_assess = TRUE` are required for LLM review. Supplying LLM options while
`seed_semantics = FALSE` warns once and does not call the provider.

## Recommended Context Bundle

For a realistic Salmon Data Package review, pass a small bundle that mixes:

1. a README, HTML export, or methods note describing the dataset,
2. a CSV, workbook, or DOCX/R Markdown data dictionary or analyst note,
3. a technical report or PDF summary if one exists.

For example:

```{r, purl = FALSE}
context_files <- c(
  "README.md",
  "methods-note.Rmd",
  "data-dictionary.xlsx",
  "technical-report.pdf"
)
```

Plain-text and CSV inputs are normally read as UTF-8. If they contain invalid
UTF-8, `metasalmon` retries Windows-1252/Latin-1 decoding before the text is
scored. R Markdown and HTML inputs are expected to be authored as UTF-8.

Each assessment records the contributing labels in `llm_context_sources`.
Ordinary files keep their base name; when two files share a base name, the label
adds parent-directory context (and, if necessary, a numeric suffix) so their
chunks and source reports remain distinct.

## One-shot `create_sdp()` Workflow

For DFO internal users, `chapi` plus the default Mistral model is the shortest path:

```{r, purl = FALSE}
library(metasalmon)

data_path <- system.file("extdata", "nuseds-fraser-coho-2023-2024.csv", package = "metasalmon")
fraser_coho <- readr::read_csv(data_path, show_col_types = FALSE)

pkg_path <- create_sdp(
  fraser_coho,
  path = "fraser-coho-2023-2024-sdp",
  dataset_id = "fraser-coho-2023-2024",
  table_id = "escapement",
  llm_assess = TRUE,
  llm_provider = "chapi",
  llm_model = "ollama2.mistral:7b",
  llm_context_files = context_files,
  check_updates = FALSE,
  overwrite = TRUE
)
```

That writes a review-ready package and uses the LLM to judge deterministic candidates during semantic seeding.

What gets written back automatically:

- accepted column-level drafts into `metadata/column_dictionary.csv` as `REVIEW: <iri>`
- accepted table observation-unit drafts into `metadata/tables.csv` as `REVIEW: <iri>` when the suggestion is still lexically compatible with the table metadata

What stays in `semantic_suggestions.csv` for manual review:

- dataset-level keyword suggestions targeting `metadata/dataset.csv`
- code-level semantic suggestions targeting `metadata/codes.csv`
- any additional shortlist evidence and `llm_*` review columns

## Full Metadata Review With `suggest_semantics()`

If you want to inspect every metadata target explicitly, start from inferred package artifacts and pass `codes`, `table_meta`, and `dataset_meta` back into `suggest_semantics()`:

```{r, purl = FALSE}
artifacts <- infer_salmon_datapackage_artifacts(
  resources = list(escapement = fraser_coho),
  dataset_id = "fraser-coho-2023-2024",
  table_id = "escapement",
  seed_semantics = FALSE
)

reviewed_dict <- suggest_semantics(
  df = artifacts$resources,
  dict = artifacts$dict,
  codes = artifacts$codes,
  table_meta = artifacts$table_meta,
  dataset_meta = artifacts$dataset_meta,
  llm_assess = TRUE,
  llm_provider = "chapi",
  llm_model = "ollama2.mistral:7b",
  llm_context_files = context_files
)

suggestions <- attr(reviewed_dict, "semantic_suggestions")
assessments <- attr(reviewed_dict, "semantic_llm_assessments")
```

Now you can filter by target file:

```{r, purl = FALSE}
suggestions[, c("target_sdp_file", "target_sdp_field", "table_id", "column_name", "code_value", "label", "iri", "llm_decision", "llm_selected")]
```

Look especially at:

- `target_sdp_file == "column_dictionary.csv"`
- `target_sdp_file == "codes.csv"`
- `target_sdp_file == "tables.csv"`
- `target_sdp_file == "dataset.csv"`

That is the clearest path when you want the LLM to help review semantics across **all** package metadata tables before you finalize anything.

## Understanding the Review Decisions

Each reviewed row carries an `llm_decision` worth scanning before you trust a draft:

- `accept` — the model picked a candidate from the deterministic shortlist; on the `create_sdp()` path it is written back as a `REVIEW:` draft.
- `review` — the model was not confident enough to pick; nothing is auto-selected and the row is left for your judgement.
- `retry_search` — the model asked for a better query; the package runs at most one bounded retry round, re-retrieves, and reassesses once.
- `reject_shortlist` — the model judged every candidate to be the wrong concept family.
- `request_new_term` — a likely ontology gap. The model can return this directly, and a `reject_shortlist` that the bounded retry cannot resolve is **escalated** to it automatically. These rows are your cue to propose a new term rather than force a near miss — see the term-request workflow (`detect_semantic_term_gaps()`, `render_ontology_term_request()`, `submit_term_request_issues()`).

The LLM never mints IRIs: `accept` only ever selects from the retrieved shortlist, and a genuine gap surfaces as `request_new_term` instead of a fabricated term.

Provider responses are handled conservatively. If one item in a batched response
is malformed, missing, or repeats a `target_key`, only the affected target is
retried individually; valid sibling assessments are preserved. If the provider
fails for the entire review but deterministic candidates are usable, the package
warns and retains that deterministic shortlist for manual review.

## Review Order

After `create_sdp()` or `suggest_semantics()`:

1. open `README-review.txt`,
2. review `metadata/column_dictionary.csv`,
3. review `metadata/tables.csv`,
4. review `metadata/dataset.csv`,
5. review `metadata/codes.csv` when present,
6. use `semantic_suggestions.csv` as the fallback evidence table.

Keep or edit every `REVIEW:` draft in the metadata CSVs directly. The `semantic_suggestions.csv` file is evidence, not the canonical package state.

## Rebuild EDH XML After Review

Once the package metadata is finalized:

```{r, purl = FALSE}
validate_salmon_datapackage(pkg_path, require_iris = TRUE)
write_edh_xml_from_sdp(pkg_path)
```

`write_edh_xml_from_sdp()` is intentionally strict. It refuses to rebuild from packages that still contain `REVIEW:` markers or unresolved dataset/table placeholders. That means the expected path is:

1. create the package,
2. review and finalize the metadata CSVs,
3. remove all `REVIEW:` prefixes,
4. run strict validation,
5. rebuild the EDH XML.

## Setup Reminder

If you have not configured the provider yet, go back to:

- [Setup and Credentials](https://salmon-data-mobilization.github.io/metasalmon/articles/setup.html)

If you want the one-shot package walkthrough first, go back to:

- [5-Minute Quickstart](https://salmon-data-mobilization.github.io/metasalmon/articles/metasalmon.html)
