---
title: "5-Minute Quickstart"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{5-Minute Quickstart}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE, purl = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Before You Start

Complete the one-time [Setup and Credentials](https://salmon-data-mobilization.github.io/metasalmon/articles/setup.html) guide first if you want to make sure GitHub installs work cleanly and any optional LLM provider is configured before you begin.

## Installation

```{r, purl = FALSE}
install.packages("remotes")
remotes::install_github("salmon-data-mobilization/metasalmon")
```

## One-shot Workflow

Load the built-in Fraser Coho 2023-2024 example (173 rows) and create a review-ready Salmon Data Package in one call.

```{r, purl = FALSE}
library(metasalmon)

data_path <- system.file("extdata", "nuseds-fraser-coho-2023-2024.csv", package = "metasalmon")
fraser_coho <- readr::read_csv(data_path, show_col_types = FALSE)

pkg_path <- create_sdp(
  fraser_coho,
  path = "fraser-coho-2023-2024-sdp",
  dataset_id = "fraser-coho-2023-2024",
  table_id = "escapement",
  check_updates = FALSE,
  overwrite = TRUE
)

pkg_path
list.files(pkg_path, recursive = TRUE)
```

If `path` is omitted, `create_sdp()` writes to your working directory using a default folder name like `fraser-coho-2023-2024-sdp`. In interactive use it can also mention when a newer `metasalmon` release is available; set `check_updates = FALSE` to skip that check.

This quickstart uses the bundled `nuseds-fraser-coho-2023-2024.csv` example, a 173-row Fraser coho slice derived from the official Open Government Canada Fraser and BC Interior workbook.

Open Government Canada record: <https://open.canada.ca/data/en/dataset/c48669a3-045b-400d-b730-48aafe8c5ee6>

See `example-data-README.md` in the bundled `extdata` folder for provenance and licensing. The package also ships a matching starter dictionary at `nuseds-fraser-coho-2023-2024-column_dictionary.csv` if you want a ready-made context file for the package-native LLM review path.

If you also need the DFO Enterprise Data Hub / GeoNetwork XML, use the one-shot path:

```{r, purl = FALSE}
pkg_path <- create_sdp(
  fraser_coho,
  path = "fraser-coho-2023-2024-sdp",
  dataset_id = "fraser-coho-2023-2024",
  table_id = "escapement",
  include_edh_xml = TRUE,
  check_updates = FALSE,
  overwrite = TRUE
)
```

That writes the HNAP-aware EDH XML to `metadata/metadata-edh-hnap.xml`. Because
`create_sdp()` produces review-ready metadata, this first XML is a **draft** when
`REVIEW:` or `MISSING` markers remain, and the package warns accordingly. Finalize
the metadata CSVs and rebuild the submission-ready XML with
`write_edh_xml_from_sdp(pkg_path)`. If you are working from an already finalized
`dataset.csv` row instead of a one-shot package build, call
`edh_build_hnap_xml()` directly.

## Optional LLM Review Later

The base quickstart does **not** require an API key. If you want `llm_assess = TRUE`, finish the one-time [Setup and Credentials](https://salmon-data-mobilization.github.io/metasalmon/articles/setup.html) guide first, then rerun the `create_sdp()` call with your chosen provider.

When you want the LLM to use supporting README notes, CSV dictionaries, Excel workbooks, or PDF reports as context, continue with:

- [LLM Review With Context Files](https://salmon-data-mobilization.github.io/metasalmon/articles/llm-context-review.html)

## Review In R

`create_sdp()` gives you a package that is **review-ready, not finished**. Two
calls take it the rest of the way, and both print the exact call for the next
step — you paste it, and the script you end up with is the record of what you
decided.

**1. Decide the semantic IRIs.** `review_semantics()` queues every unfilled
slot with its ranked candidates and their definitions:

```{r, purl = FALSE}
review <- review_semantics(pkg_path)
review
```

```
── escapement · spawner_count · variable ──────────────────────────
   field:   column_dictionary.csv · term_iri
   current: REVIEW: https://w3id.org/smn/SpawnerAbundance

   [1] Spawner Abundance   smn   score 4.9
       The number of mature salmon returning to spawn in a stream.
       https://w3id.org/smn/SpawnerAbundance
       review <- accept_suggestion(review, "spawner_count", "variable", rank = 1)
```

Paste the line for the candidate you want, or the printed
`reject_suggestion(..., reason = "…")` when none fits — the reason is written
into the package, so the next reader can see why. Then write the decisions
back:

```{r, purl = FALSE}
review <- accept_suggestion(review, "spawner_count", "variable", rank = 1)
apply_sdp_semantics(pkg_path, review)
```

`apply_sdp_semantics()` is surgical and safe to re-run: it strips the `REVIEW:`
prefix from the fields you decided, leaves the ones you did not alone, and does
not touch a single byte of your data CSVs. Decisions persist, so you can stop
and pick the review up tomorrow without being asked the same questions again.

**2. Fill in the free text and anything with no candidates.**
`review_metadata()` reads the package against the rules that decide strict
validation rather than against the shortlist, so it sees something the semantic
review structurally cannot: a required field that **no candidate was ever found
for**.

```{r, purl = FALSE}
review_metadata(pkg_path)
```

```
── dataset.csv ────────────────────────────────────────────────────
   creator: placeholder text, refused by strict validation
      MISSING METADATA: add creator, team, or originating program.

   set_sdp_dataset(pkg_path,
     creator = "<add creator, team, or originating program>"
   )
```

Replace each `<…>` with the real value and paste. (Pasting one unedited is
refused — a package whose `creator` reads `<add creator…>` would pass
validation while saying nothing.) The four setters are `set_sdp_dataset()`,
`set_sdp_table()`, `set_sdp_column()` and `set_sdp_code()`; each keeps
`datapackage.json` in step in the same write.

When `review_metadata()` reports nothing, strict validation passes:

```{r, purl = FALSE}
validate_salmon_datapackage(pkg_path, require_iris = TRUE)
```

Editing `metadata/*.csv` in a spreadsheet still works and is still supported.
It is the fallback rather than the recommended path, for one reason: a
spreadsheet edit leaves no record of **why** a term was chosen, and that record
is the most valuable thing the review produces. `README-review.txt` in the
package folder carries the same checklist.

The path to `metadata/column_dictionary.csv` is also a reasonable
`llm_context_files` input when you want the LLM review step to reason from the
package itself instead of a separate methods note. Pass the path string, not a
tibble created by reading the CSV, and set `llm_assess = TRUE`; context alone does
not enable an LLM call.

`create_sdp()` seeds semantic suggestions by default and auto-fills top-ranked compatible drafts directly into blank semantic fields in the metadata CSVs. That includes variable, property, entity, and unit IRIs in `metadata/column_dictionary.csv` plus strong table-level observation-unit drafts in `metadata/tables.csv`, using `observation_unit`/`description` when available and otherwise falling back to `table_label`/`table_id`. Constraint and statistical-modifier candidates are held back unless the column's own name, label, or description carries the evidence for them, and under LLM review (`llm_assess = TRUE`) they are never auto-filled at all. Any auto-applied semantic IRI draft is written back as `REVIEW: <iri>` so you still confirm it manually. It does not overwrite existing non-empty semantic values. Code-level suggestions default to factor and low-cardinality character source columns; use `semantic_code_scope = "all"` if you want broader code-level seeding.

When LLM review is enabled, the six measurement roles — variable, property,
entity, unit, constraint, and statistical modifier — are judged together and
then checked by deterministic validators. A validator can downgrade an
unsupported acceptance to manual review, but it never substitutes or invents a
term. Omit `semantic_sources` for role-aware defaults; an explicitly supplied
vector is a strict allowlist through the single retry round.

The inferred metadata includes `MISSING DESCRIPTION:` and `MISSING METADATA:` placeholders for required fields, so the package tells you what it is still missing. `review_metadata()` lists exactly those fields and prints the call that fills each one. Expect `validate_salmon_datapackage()` to **warn about every placeholder it finds** (since 0.2.6), including on a package you created seconds ago. That warning on a fresh package is normal: it is the to-do list, not a defect. The `metadata/*.csv` files are the canonical package metadata; `datapackage.json` is a derived export for interoperability.

## How To Decide If `term_iri` Is Correct

Use plain-language checks for each measurement column:

1. Does the suggested label describe exactly what the column measures?
2. Does the definition match your intent (not just a similar word)?
3. Is the scope right (for example species-level vs population-level)?
4. Is the unit consistent with your values and `unit_iri`?

Keep the IRI only when all checks pass.

Replace it when the term is close but not exact.

Remove it (leave blank) when no candidate is reliable yet.

When the top auto-applied suggestion is wrong, `review_semantics()` prints the whole shortlist with each candidate's definition; accept a different one with the `rank =` it shows. When the right term exists but retrieval never surfaced it, pass it directly: `accept_suggestion(review, "spawner_count", "variable", iri = "…")`, or `set_sdp_column(pkg_path, "spawner_count", term_iri = "…")` for a slot with no shortlist at all.

If no candidate fits, request a new term instead of forcing a bad match:

- shared cross-organization/domain terms -> <https://github.com/salmon-data-mobilization/salmon-domain-ontology/issues/new/choose>
- DFO-specific policy/operations terms -> <https://github.com/dfo-pacific-science/dfo-salmon-ontology/issues/new/choose>

## Finalize

`apply_sdp_semantics()` and the `set_sdp_*()` setters write straight into the
package, so there is nothing to save or reload. If you did edit the CSVs in a
spreadsheet, save them back as CSV and reload:

```{r, purl = FALSE}
pkg <- read_salmon_datapackage(pkg_path)
validate_salmon_datapackage(pkg_path, require_iris = FALSE)
```

That review-state validation is the first follow-up check. For the full post-review workflow — reloading the package, detecting unresolved semantic gaps, deciding shared salmon-domain vs DFO-specific routing, drafting term requests, rebuilding EDH XML if needed, and only then running strict final validation — continue with:

- [After Excel Review: Finalize and Publish Your Package](https://salmon-data-mobilization.github.io/metasalmon/articles/post-review-package-publication.html)

For a staged, fully explicit workflow where you assemble metadata tables manually instead of continuing from a reviewed `create_sdp()` package, use:

- [Publishing Data Packages](https://salmon-data-mobilization.github.io/metasalmon/articles/data-dictionary-publication.html)

Two more guides are worth knowing about before you package real data:

- [Tidy Data for Salmon Data Packages](https://salmon-data-mobilization.github.io/metasalmon/articles/tidy-data-for-sdp.html) — what the primary-key and untidy-column-name checks mean, and how to reshape a year-per-column table.
- [Migrating to SDP 0.3.0](https://salmon-data-mobilization.github.io/metasalmon/articles/migrating-to-sdp-0-3-0.html) — required reading if you have a package built with metasalmon 0.2.x, since methods left the column dictionary.
