---
title: "Migrating to SDP 0.3.0: Where Methods Live Now"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Migrating to SDP 0.3.0: Where Methods Live Now}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE, purl = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Overview

metasalmon 0.3.0 implements version `sdp-0.3.0` of the Salmon Data Package
specification, and that version moved **methods** out of the column dictionary.
If you have a package built with metasalmon 0.2.x, this guide takes you through
the change: why it happened, where each piece of your method metadata belongs
now, and how to run the migration tool on your existing package.

You do not need to have followed the specification discussion. You do need to
be able to open your package's `metadata/` CSVs.

The migration itself needs no internet connection and no LLM. The complete
worked example below runs offline in a few seconds, and the console output shown
beside it is real output from that example. Code chunks are not executed when
this page is built, so you can copy them and run them yourself. Two later
sections — term search and term requests — do reach the network, and are marked
where they appear.

If you are starting a brand-new package, you do not need this guide at all —
`create_sdp()` already writes the new shape. Read the section
[Where a method belongs now](#where-a-method-belongs-now) for the concepts, and
skip the migration sections.

## What Changed, in One Paragraph

In sdp-0.2.0, a measurement column in `metadata/column_dictionary.csv` could
carry a `method_iri`, and an optional `metadata/methods.csv` registry held
labels, descriptions, versions, and citations for those methods. In sdp-0.3.0
the dictionary's `method_iri` column is **gone**, the registry file is **gone**,
and a method now lives in one of three places depending on what it actually
describes. In exchange, the dictionary gained a new column,
`statistical_modifier_iri`, for the one kind of "how" that really is part of
what a column measures.

## Why Methods Moved

A dictionary row answers the question *what is this column?* A method answers a
different question: *how was this value produced?* Those are not the same
question, and putting the answer to the second one inside the answer to the
first one let them blur together.

Two examples make the difference concrete.

**A weir count and an aerial count are the same variable.** Both columns hold
the number of adult spawners in a stream. If you counted at a weir and your
colleague counted from a float plane, you measured the same thing with different
instruments. Your data should link to the *same* term for "spawner count", so
that anyone combining your datasets gets one variable, not two. Recording the
counting method inside the variable's identity quietly split it in half.

**A mean weight and a maximum weight are different variables.** Here the summary
really is part of what the column *is*. A column of mean weights and a column of
maximum weights cannot be pooled, compared, or averaged as though they were the
same measurement. The aggregation belongs to the variable's identity.

The old `method_iri` column could not tell those two cases apart, so sdp-0.3.0
splits them. Aggregations get their own dictionary column,
`statistical_modifier_iri`. Everything else — the gear, the survey design, the
laboratory procedure, the field protocol — moves out of the dictionary to the
level where it is actually constant.

## Where a Method Belongs Now {#where-a-method-belongs-now}

There are exactly three placements. Pick by asking **what the method is constant
across**.

| If the method is... | It goes in | Field |
|---|---|---|
| The same for every row of a table | `metadata/tables.csv` | `method_iri` |
| A published procedure document you want to cite | `metadata/tables.csv` or `metadata/dataset.csv` | `protocol_iri`, `protocol_citation` |
| Different from row to row | your data, as a code column | `metadata/codes.csv` `term_iri` |

### Placement 1: One Method for the Whole Table

Use this when every row of a table was produced the same way. A table of weir
counts is the classic case: the weir is the method, and it applies to the whole
table.

```{r}
tables <- readr::read_csv("weir-counts-sdp/metadata/tables.csv", na = "")

tables$method_iri[tables$table_id == "escapement"] <-
  "https://example.org/methods/weir-count"

readr::write_csv(tables, "weir-counts-sdp/metadata/tables.csv", na = "")
```

The value must be an **absolute IRI**. Since 0.3.0, a table or dataset
method/protocol field holding something like `methods/weir-count` is reported by
`validate_salmon_datapackage()` rather than silently accepted.

A table-level `method_iri` is a genuine vocabulary term, so it carries the same
obligations as any other reviewed IRI when you export or publish: EML export
requires an accepted semantic-review ledger row for it, and requires it to
appear in the reviewed vocabulary snapshot. (A `protocol_iri` is a citation, not
a vocabulary term, and is not gated that way.) If you do not have a term for
your method yet, see
[What to Do With the Old Registry](#what-to-do-with-the-old-registry) below.

### Placement 2: Cite a Protocol

Use this when what you want to record is a **document** — a published field
manual, a standard operating procedure, a methods paper — rather than a
vocabulary concept. Protocols are citations, not vocabulary terms, so they are
not checked against the ontology.

Put them on `tables.csv` when the protocol governs one table, and on
`dataset.csv` when it governs the whole dataset.

```{r}
tables$protocol_iri[tables$table_id == "escapement"] <-
  "https://example.org/protocols/weir-2019"
tables$protocol_citation[tables$table_id == "escapement"] <-
  "Salmon Program. 2019. Weir enumeration protocol v3.1."
```

`protocol_iri` is where the document lives; `protocol_citation` is the
human-readable reference. You can supply either or both, and you can use a
protocol *alongside* a table-level `method_iri` — the concept says what
procedure was used, the citation says where it is written down.

### Placement 3: The Method Varies by Row

Use this when different rows of the same table were produced differently — a
long-running escapement series where some years were flown and some were
counted at a fence, for example. A method that varies by row is **data**, not
metadata, so it belongs in a column of your data file.

Add the column to your data, describe it in the dictionary as a code column,
and then give each code a definition in `metadata/codes.csv`, pointing
`term_iri` at the shared-vocabulary procedure it means.

```{r}
# 1. The method becomes a column in the data itself.
escapement <- readr::read_csv("weir-counts-sdp/data/escapement.csv", na = "")
escapement$enumeration_method <- c("weir", "aerial", "weir")
readr::write_csv(escapement, "weir-counts-sdp/data/escapement.csv", na = "")

# 2. Describe the new column in the dictionary, like any other column.
dictionary <- readr::read_csv(
  "weir-counts-sdp/metadata/column_dictionary.csv", na = ""
)
dictionary <- dplyr::bind_rows(dictionary, tibble::tibble(
  dataset_id         = "weir-counts",
  table_id           = "escapement",
  column_name        = "enumeration_method",
  column_label       = "Enumeration method",
  column_description = "How the spawners in this row were counted.",
  column_role        = "categorical",
  value_type         = "string"
))
readr::write_csv(
  dictionary, "weir-counts-sdp/metadata/column_dictionary.csv", na = ""
)

# 3. Each code value resolves to a shared procedure concept.
codes <- tibble::tibble(
  dataset_id       = "weir-counts",
  table_id         = "escapement",
  column_name      = "enumeration_method",
  code_value       = c("weir", "aerial"),
  code_label       = c("Weir count", "Aerial survey count"),
  code_description = c(
    "Complete enumeration of adults passing a fixed weir.",
    "Visual count of adults from a fixed-wing aircraft or helicopter."
  ),
  vocabulary_iri   = "https://example.org/methods/",
  term_iri         = c(
    "https://example.org/methods/weir-count",
    "https://example.org/methods/aerial-survey-count"
  ),
  term_type        = "sosa:Procedure"
)

readr::write_csv(codes, "weir-counts-sdp/metadata/codes.csv", na = "")
```

This is the placement that the migration tool cannot make for you, because it
requires adding a column to your data — a change only you can make correctly.

If your method values come from NuSEDS, the bundled crosswalks give you a
starting map from the legacy labels to canonical method families:

```{r}
nuseds_enumeration_method_crosswalk()
nuseds_estimate_method_crosswalk()
```

## The New `statistical_modifier_iri` Column

This is the piece that moved *into* the dictionary rather than out of it.

Fill `statistical_modifier_iri` when the column holds an **aggregation or
summary** rather than a single observation: a mean, a maximum, a minimum, a
total, a peak. That summary is part of the variable's identity, which is why it
lives beside `property_iri`, `entity_iri`, and `constraint_iri` as the fifth
I-ADOPT component.

Leave it blank for plain measurements. A column of individual fish weights has
no statistical modifier; a column of *mean* fish weights per sample does.

The semantic pipeline is deliberately conservative here. It proposes a
statistical-modifier candidate only when the column's name, label, or
description actually names an aggregation — the words it looks for are *total*,
*cumulative*, *sum*, *mean*, *average*, *max*/*maximum*, *min*/*minimum*, and
*peak*. So `mean_fork_length` gets a suggestion and `fork_length` does not, and
ordinary measurement columns do not gain review work they do not need.

When you do want candidates for the slot, `sources_for_role()` knows the role.
`sources_for_role()` is a local lookup; `find_terms()` queries vocabulary
services and needs a network connection.

```{r}
sources_for_role("statistical_modifier")

find_terms("mean", role = "statistical_modifier")
```

A quick way to tell the two apart: if you changed how you counted and the
numbers are still the same kind of number, that is a **method**. If you changed
it and the numbers now mean something different, that is a **statistical
modifier**.

## Migrating an Existing Package

`migrate_sdp_methods()` does the mechanical part of the move and **stops and
reports** on anything that needs your judgement. It never guesses. The rewrite
is atomic: if the migration stops, nothing on disk has changed, so you can fix
the problem and run it again.

### Always Dry-Run First

```{r}
library(metasalmon)

report <- migrate_sdp_methods("weir-counts-sdp", dry_run = TRUE)
```

`dry_run = TRUE` reports exactly what a real run would do without touching a
single file. Read the report, decide whether you agree with it, and only then
run it for real.

### A Complete Worked Example

This builds a small sdp-0.2.0-shaped package in a temporary directory, migrates
it, and shows the result. It runs offline in a few seconds and leaves nothing
behind on your machine.

First, write the legacy package: three columns, one measurement column carrying
a dictionary `method_iri`, and a `metadata/methods.csv` registry describing that
method.

```{r}
library(metasalmon)

legacy_path <- file.path(tempdir(), "weir-counts-sdp")
dir.create(file.path(legacy_path, "data"), recursive = TRUE)
dir.create(file.path(legacy_path, "metadata"), recursive = TRUE)

readr::write_csv(
  tibble::tibble(
    stream_id     = c("BEAR", "BEAR", "COHO"),
    survey_year   = c(2023L, 2024L, 2023L),
    spawner_count = c(412L, 388L, 1204L)
  ),
  file.path(legacy_path, "data", "escapement.csv"),
  na = ""
)

readr::write_csv(
  tibble::tibble(
    dataset_id   = "weir-counts",
    title        = "Weir counts",
    description  = "Adult spawner counts at fixed weirs.",
    spec_version = "sdp-0.2.0"
  ),
  file.path(legacy_path, "metadata", "dataset.csv"),
  na = ""
)

readr::write_csv(
  tibble::tibble(
    dataset_id  = "weir-counts",
    table_id    = "escapement",
    file_name   = "data/escapement.csv",
    table_label = "Escapement",
    primary_key = "stream_id,survey_year"
  ),
  file.path(legacy_path, "metadata", "tables.csv"),
  na = ""
)

readr::write_csv(
  tibble::tibble(
    dataset_id         = "weir-counts",
    table_id           = "escapement",
    column_name        = c("stream_id", "survey_year", "spawner_count"),
    column_label       = c("Stream", "Survey year", "Spawner count"),
    column_description = c("Stream code.", "Year surveyed.", "Adult spawners counted."),
    column_role        = c("identifier", "identifier", "measurement"),
    value_type         = c("string", "integer", "integer"),
    # The sdp-0.2.0 placement this migration removes:
    method_iri         = c("", "", "https://example.org/methods/weir-count")
  ),
  file.path(legacy_path, "metadata", "column_dictionary.csv"),
  na = ""
)

readr::write_csv(
  tibble::tibble(
    dataset_id         = "weir-counts",
    method_iri         = "https://example.org/methods/weir-count",
    method_label       = "Weir count",
    method_description = "Complete enumeration of adults passing a fixed weir.",
    method_version     = "3.1",
    protocol_iri       = "https://example.org/protocols/weir-2019",
    citation           = "Salmon Program. 2019. Weir enumeration protocol v3.1."
  ),
  file.path(legacy_path, "metadata", "methods.csv"),
  na = ""
)
```

Now preview the migration:

```{r}
report <- migrate_sdp_methods(legacy_path, dry_run = TRUE)
```

```
Table-level method placements:
✔ escapement -> https://example.org/methods/weir-count (from spawner_count)
'metadata/methods.csv' is removed by this migration. Its labels and
descriptions belong in the shared vocabulary; its version and citation belong
beside protocol_iri:
• https://example.org/methods/weir-count (Weir count)
ℹ Request missing vocabulary terms through the ontology's shared-term admission
  policy, and copy any registry citation into protocol_citation.
Dry run: no files were changed.
```

The report is also returned, invisibly, as a list with three parts: `tables`
(the placements it will apply), `dropped_review` (unresolved `REVIEW:` bindings
it will drop), and `registry` (the rows of the old `methods.csv`, so you can
relocate their content).

```{r}
report$tables
```

```
# A tibble: 1 × 3
  table_id   method_iri                             columns
  <chr>      <chr>                                  <chr>
1 escapement https://example.org/methods/weir-count spawner_count
```

```{r}
report$registry[c("method_label", "method_version", "citation")]
```

```
# A tibble: 1 × 3
  method_label method_version citation
  <chr>        <chr>          <chr>
1 Weir count   3.1            Salmon Program. 2019. Weir enumeration protocol v…
```

The dry run agrees with what we intended, so run it for real:

```{r}
report <- migrate_sdp_methods(legacy_path)
```

The same report prints, followed by:

```
✔ Migration complete.
ℹ Run `validate_salmon_datapackage("...")` to confirm the package.
```

Check the result. The method is now a property of the table:

```{r}
tables <- readr::read_csv(
  file.path(legacy_path, "metadata", "tables.csv"),
  show_col_types = FALSE, na = ""
)
tables[c("table_id", "method_iri", "protocol_iri", "protocol_citation")]
```

```
# A tibble: 1 × 4
  table_id   method_iri                           protocol_iri protocol_citation
  <chr>      <chr>                                <lgl>        <lgl>
1 escapement https://example.org/methods/weir-co… NA           NA
```

The dictionary has lost `method_iri` and gained `statistical_modifier_iri`:

```{r}
names(readr::read_csv(
  file.path(legacy_path, "metadata", "column_dictionary.csv"),
  show_col_types = FALSE, na = ""
))
```

```
 [1] "dataset_id"               "table_id"
 [3] "column_name"              "column_label"
 [5] "column_description"       "column_role"
 [7] "value_type"               "required"
 [9] "unit_label"               "unit_iri"
[11] "term_iri"                 "term_type"
[13] "property_iri"             "entity_iri"
[15] "constraint_iri"           "statistical_modifier_iri"
```

And the registry is gone:

```{r}
file.exists(file.path(legacy_path, "metadata", "methods.csv"))
#> [1] FALSE
```

Notice what the migration did *not* do: it left `protocol_iri` and
`protocol_citation` empty, even though the old registry had both. That is
deliberate. Deciding that a registry citation describes the table rather than
the dataset is a judgement call, so the migration hands you the values in
`report$registry` and lets you place them. Doing that is the next section.

The migration also updates `datapackage.json` — it drops the `sdp_methods`
resource, removes any legacy `iAdopt:methodIri` custom keys from the field
definitions, and stamps the new profile and spec version — and sets
`dataset.csv`'s `spec_version` to `sdp-0.3.0`.

### Finally, Validate

```{r}
validate_salmon_datapackage(legacy_path, require_iris = FALSE)
```

Run this after every migration. It is also worth knowing that a package still
carrying a `metadata/methods.csv` is now refused by the surfaces that consume
method metadata — EML export, KNB publication, and observation-structure
validation — with an error pointing back at `migrate_sdp_methods()`.

## When the Migration Stops

`migrate_sdp_methods()` stops rather than guessing whenever the old metadata
does not determine a single correct answer. Each stop leaves your package
exactly as it was. Here is each case and what to do about it.

### Columns of One Table Bound to Different Methods

```
Error in migrate_sdp_methods(path, dry_run = TRUE) :
  Method migration stopped: measurement columns disagree about their
table's method.
✖ Table escapement: https://example.org/methods/electrofishing (fry_density) vs
  https://example.org/methods/weir-count (spawner_count)
ℹ Split the table, cite a protocol instead, or move the method into the data as
  a code column, then re-run.
ℹ See the methods section of the SDP specification for the three placements.
```

The tool cannot promote a method to the table when the table's columns disagree
about what it is. You have three ways out, and which one is right depends on
your data:

1. **Split the table.** If `fry_density` and `spawner_count` really are separate
   observational units produced by separate field programs, they are two tables.
   This is usually the honest answer, and it also makes the primary key easier
   to declare.
2. **Cite a protocol instead.** If both columns come from one field programme
   documented in one manual, drop the per-column methods and put the manual in
   `protocol_iri` / `protocol_citation` on the table.
3. **Move the method into the data.** If the difference is real and varies, add
   an enumeration-method column as in [Placement 3](#placement-3-the-method-varies-by-row).

A related message appears when a method is bound to *some* measurement columns
of a table but not all of them:

```
✖ Table escapement: https://example.org/methods/weir-count is bound to only
  some measurement columns; fry_density carries no resolved method binding.
```

Promotion claims the method for the **whole** table, so silence from a column is
treated as disagreement rather than consent. Either bind the missing columns to
the same method, or take one of the three options above.

### Bindings Naming an Undeclared Table

```
Error: Method bindings name tables that 'metadata/tables.csv' does not declare.
✖ escapement
ℹ Fix the table identifiers in the legacy metadata, then re-run.
```

A dictionary row claims `table_id = "escapement"`, but `tables.csv` has no such
table — usually a typo or a table that was renamed on one side only. Fix the
identifier so both files agree, then re-run. A closely related stop fires when a
binding has no `table_id` or no `column_name` at all:

```
Error: Method bindings without a table and column cannot be migrated.
```

One practical note: the undeclared-table check runs when the migration writes,
so a `dry_run = TRUE` preview does not surface it. The write is atomic, so a
real run that hits this stop still leaves your package untouched — you simply
find out at the real run rather than at the preview.

You may also see two carriers disagreeing about one column, if your package has
both a dictionary `method_iri` and a legacy `iAdopt:methodIri` custom key in
`datapackage.json`:

```
Error: Method migration stopped: two carriers disagree about one column's method.
```

The tool refuses to pick a winner, because picking one would erase the other
from your package. Decide which is correct, make both files say it, and re-run.

### Unresolved `REVIEW:` Values Are Dropped

```
Unresolved `REVIEW:` method bindings dropped (resolve them via term search
before publishing):
✖ escapement.fry_density = REVIEW: https://example.org/methods/electrofishing
✖ escapement.spawner_count = REVIEW: https://example.org/methods/weir-count
```

A `REVIEW:` prefix means a suggestion nobody confirmed. The migration does not
carry unconfirmed suggestions into the new shape — that would launder a draft
into a fact — so it drops them and tells you which ones it dropped.

This is not an error, and the migration continues. But note what it means: those
columns now have **no** method recorded anywhere. If the method matters, resolve
the term before or after migrating (`find_terms()`, or the term-request workflow
below) and then add it in the right placement by hand.

Watch for one interaction: dropping a `REVIEW:` binding can turn a table that
looked unanimous into one where only some measurement columns carry a method,
which then produces the "bound to only some measurement columns" stop above.
Resolving the `REVIEW:` value is the cleanest fix.

## What to Do With the Old Registry {#what-to-do-with-the-old-registry}

The old `metadata/methods.csv` had seven columns. They do not all go to the same
place, and the migration reports them rather than relocating them, because two
of them need a decision and three of them need to leave your package entirely.

| Old registry column | Where it goes now |
|---|---|
| `method_iri` | `tables.csv$method_iri`, or `codes.csv$term_iri` — the migration places this |
| `method_label` | The shared vocabulary — request the term |
| `method_description` | The shared vocabulary — request the term |
| `method_version` | `protocol_citation` |
| `protocol_iri` | `protocol_iri` on `tables.csv` or `dataset.csv` |
| `citation` | `protocol_citation` on `tables.csv` or `dataset.csv` |
| `dataset_id` | Nothing — it was only there to key the registry |

### Labels and Descriptions Belong in the Vocabulary

This is the reasoning behind removing the registry. A label and a definition for
"weir count" are the same in every package that uses weir counts. Copying them
into each package's `methods.csv` meant every package restated the vocabulary,
inconsistently, with no way to correct all the copies at once. The IRI is the
link; the definition lives at the other end of it.

So if your registry described a method that has no shared term yet, the fix is
to **request the term** rather than to keep a local copy of the definition. The
package has a workflow for exactly this. It reaches GitHub, so it needs a
network connection and the credentials from the
[Setup and Credentials](setup.html) guide.

```{r}
reviewed_dict <- read_salmon_datapackage(legacy_path)$dictionary

gaps <- detect_semantic_term_gaps(reviewed_dict)

requests <- render_ontology_term_request(gaps, scope = "auto", ask = FALSE)

# Review every draft, then preview the issues without submitting them:
requests |>
  dplyr::filter(request_scope %in% c("smn", "gcdfo")) |>
  submit_term_request_issues(dry_run = TRUE)
```

Your old `method_label` becomes the proposed label and your `method_description`
becomes the proposed definition, which is usually most of a good term request
already. Route it by scope: terms that other organizations would also use go to
the shared salmon-domain ontology, and DFO-specific operational terms go to the
DFO salmon ontology.

- shared cross-organization/domain terms ->
  <https://github.com/salmon-data-mobilization/salmon-domain-ontology/issues/new/choose>
- DFO-specific policy/operations terms ->
  <https://github.com/dfo-pacific-science/dfo-salmon-ontology/issues/new/choose>

For the full term-request walkthrough, see
[After Excel Review: Finalize and Publish Your Package](post-review-package-publication.html)
and [Linking to Standard Vocabularies](reusing-standards-salmon-data-terms.html).

### Versions and Citations Become `protocol_citation`

A `method_version` such as `3.1` and a `citation` such as
`Salmon Program. 2019. Weir enumeration protocol v3.1.` are describing a
**document**, not a concept. Version 3.1 of your weir protocol and version 2.0
are the same procedure, differently written down — which is exactly what
`protocol_citation` is for.

Take them from `report$registry` and place them by hand, deciding as you go
whether the protocol governs one table or the whole dataset:

```{r}
report <- migrate_sdp_methods("weir-counts-sdp", dry_run = TRUE)

report$registry[c("method_version", "protocol_iri", "citation")]

# Then, after the real migration, place them where they belong:
tables <- readr::read_csv("weir-counts-sdp/metadata/tables.csv", na = "")
tables$protocol_iri[tables$table_id == "escapement"] <-
  "https://example.org/protocols/weir-2019"
tables$protocol_citation[tables$table_id == "escapement"] <-
  "Salmon Program. 2019. Weir enumeration protocol v3.1."
readr::write_csv(tables, "weir-counts-sdp/metadata/tables.csv", na = "")
```

If the same protocol governs everything in the dataset, put it on
`metadata/dataset.csv` instead — the fields have the same names there.

## Migration Checklist

1. Back up the package, or make sure it is committed to version control.
2. Run `migrate_sdp_methods(path, dry_run = TRUE)` and read the whole report.
3. Resolve any stop: split a table, cite a protocol, or move the method into the
   data as a code column.
4. Resolve or accept the dropped `REVIEW:` bindings.
5. Run `migrate_sdp_methods(path)` for real.
6. Place `protocol_iri` and `protocol_citation` from the registry report.
7. Fill `statistical_modifier_iri` for any aggregated columns.
8. Request shared terms for method labels and descriptions that have no IRI yet.
9. Run `validate_salmon_datapackage(path)`.

## Related Reading

- [Tidy Data for Salmon Data Packages](tidy-data-for-sdp.html) — the tidy
  foundations the method model depends on. "Is the method constant within this
  table?" is only a sound question when the table is a coherent observational
  unit.
- [Glossary of Terms](glossary.html) — the I-ADOPT components, including the
  statistical modifier.
- [Linking to Standard Vocabularies](reusing-standards-salmon-data-terms.html) —
  finding and requesting terms.
- The `NEWS.md` entry for 0.3.0 is the authoritative statement of the change.
