---
title: "After Excel Review: Finalize and Publish Your Package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{After Excel Review: Finalize and Publish Your Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Overview

This guide starts **after** you already ran `create_sdp()`, opened the package in
Excel (or another spreadsheet editor), and saved your reviewed metadata back to
CSV.

If you have not created the package yet, start with the
[5-Minute Quickstart](metasalmon.html).

If you want to assemble `dataset.csv`, `tables.csv`, and
`column_dictionary.csv` manually instead of continuing from a reviewed package,
use [Publishing Data Packages](data-dictionary-publication.html).

This is the post-review path:

1. reload the reviewed package from disk;
2. validate the reviewed metadata and data structure;
3. re-run semantic suggestions only for fields that are still unresolved;
4. detect likely ontology gaps;
5. decide whether each missing term belongs in the shared salmon-domain
   ontology or the DFO-specific bucket;
6. produce a concrete list of term requests to file;
7. rebuild EDH XML if needed; and
8. run strict final validation before publishing.

## 1) Reload the reviewed package

```{r}
library(metasalmon)

pkg_path <- "fraser-coho-2023-2024-sdp"
pkg <- read_salmon_datapackage(pkg_path)

names(pkg$resources)
pkg$tables
```

`read_salmon_datapackage()` reloads the package from the canonical
`metadata/*.csv` files, so you are checking the same metadata you just reviewed
in Excel.

## 2) Run a review-state validation pass

Start with the package-level validator in **non-strict** mode:

```{r}
review_check <- validate_salmon_datapackage(pkg_path, require_iris = FALSE)
```

This catches package/data mismatches now, before you worry about final ontology
coverage.

If you also want the semantic details separated out, inspect the reloaded
package directly:

```{r}
review_semantics <- validate_semantics(pkg$dictionary, require_iris = FALSE)

review_semantics$issues
review_semantics$missing_terms
```

Use this pass to answer two questions:

- **Did the Excel edits break the package structure?**
- **Which measurement rows still lack a final semantic term?**

Do **not** switch to `require_iris = TRUE` yet unless you believe the package is
fully finalized. Strict validation is the last gate, not the first one.

## 3) Re-run semantic suggestions for only the unresolved pieces

The package-root `semantic_suggestions.csv` file is the original evidence trail
from `create_sdp()`. After review, it is usually more useful to re-run
`suggest_semantics()` against the **current** package state so you only inspect
what is still unresolved.

```{r}
reviewed_dict <- suggest_semantics(
  df = pkg$resources,
  dict = pkg$dictionary,
  codes = pkg$codes,
  table_meta = pkg$tables,
  dataset_meta = pkg$dataset
)

remaining_suggestions <- attr(reviewed_dict, "semantic_suggestions")

remaining_suggestions |>
  dplyr::select(
    target_scope,
    target_row_key,
    target_sdp_field,
    dictionary_role,
    search_query,
    label,
    source,
    score
  )
```

That usually produces a much shorter shortlist than the original package export,
because rows you already resolved in Excel are no longer the target.

## 4) Detect likely ontology gaps

Now detect the places where `metasalmon` still cannot find a shared SMN match,
but did find useful fallback candidates elsewhere:

```{r}
gaps <- detect_semantic_term_gaps(reviewed_dict)

gaps |>
  dplyr::select(
    target_scope,
    target_row_key,
    target_sdp_field,
    dictionary_role,
    search_query,
    top_non_smn_label,
    top_non_smn_source,
    placement_recommendation,
    placement_confidence
  )
```

This is the key post-review gap table:

- rows here are still unresolved;
- `top_non_smn_*` columns show the best non-SMN evidence the package found;
- `placement_recommendation` gives a first-pass routing guess.

If you only want a narrower slice, filter by role or scope:

```{r}
gaps |>
  dplyr::filter(dictionary_role %in% c("variable", "property", "entity"))
```

## 5) Decide shared salmon-domain vs DFO-specific routing

Use this plain-English rule:

- **Shared salmon-domain ontology (`smn`)**: use this when the term describes a
  reusable salmon science concept that another program, region, or organization
  could reasonably use too.
- **DFO-specific term**: use this when the concept is clearly tied to DFO policy,
  operations, internal workflow, local identifiers, program-specific statuses,
  or other context that is not broadly reusable outside that setting.

A good default test is:

- if the term would still make sense in a non-DFO salmon dataset, lean **SMN**;
- if the term only makes sense because of a DFO or local program process, lean
  **DFO-specific**.

### How that maps to the current helper outputs

`detect_semantic_term_gaps()` and `render_ontology_term_request()` currently use
these buckets:

- `placement_recommendation == "smn"` → likely shared salmon-domain request
- `placement_recommendation == "profile"` → likely **not** shared; in practical
  DFO use, treat this as the DFO/program-specific bucket
- `placement_recommendation == "uncertain"` → make a human decision before
  drafting requests

> **Current limitation:** the non-shared bucket is still named `profile` in the
> helper API. That is the right bucket for DFO/program-specific requests in this
> workflow, but the name is more generic than the real governance choice.

## 6) Produce a concrete term-request list

First build a simple routing table you can save with the package:

```{r}
request_plan <- gaps |>
  dplyr::mutate(
    request_scope = dplyr::case_when(
      placement_recommendation == "smn" ~ "smn",
      placement_recommendation == "profile" ~ "profile",
      TRUE ~ "skip"
    ),
    governance_bucket = dplyr::case_when(
      request_scope == "smn" ~ "shared salmon-domain ontology (SMN)",
      request_scope == "profile" ~ "DFO/program-specific ontology bucket",
      TRUE ~ "needs manual routing decision"
    )
  ) |>
  dplyr::select(
    dataset_id,
    table_id,
    target_scope,
    target_row_key,
    target_sdp_field,
    dictionary_role,
    search_query,
    top_non_smn_label,
    top_non_smn_source,
    placement_recommendation,
    placement_confidence,
    governance_bucket,
    request_scope
  )

request_plan
readr::write_csv(request_plan, file.path(pkg_path, "term-request-plan.csv"))
```

That gives you a concrete list of:

- which package row still needs help;
- what the best fallback term looked like;
- where the request should probably go.

Next, render draft request text from that plan:

```{r}
requests <- render_ontology_term_request(
  gaps,
  scope = "auto",
  ask = FALSE,
  profile_name = "dfo-salmon",
  scope_overrides = request_plan$request_scope
)

requests |>
  dplyr::select(request_scope, request_title, target_row_key, search_query)
```

### Preview shared-term issue drafts

For shared salmon-domain requests, the package already has a dry-run GitHub path:

```{r}
smn_preview <- requests |>
  dplyr::filter(request_scope == "smn") |>
  submit_term_request_issues(dry_run = TRUE)

smn_preview
```

That preview is usually enough to review the issue titles/bodies before posting.

### What to do with DFO-specific rows

Rows with `request_scope == "profile"` are the DFO/program-specific cases in this
workflow. Keep them in `term-request-plan.csv` and file them through the right
DFO governance process after human review.

At the moment, `submit_term_request_issues()` is best treated as the built-in
path for **shared SMN** requests, not as a one-click publisher for DFO-specific
rows.

## 7) Finalize metadata and rebuild EDH XML if needed

Once the metadata is final and every surviving IRI is deliberate, regenerate EDH
XML if your publication path needs it:

```{r}
write_edh_xml_from_sdp(pkg_path)
```

This is the reviewed-package wrapper around `edh_build_hnap_xml()`. It refuses
to rebuild while obvious review-state markers still exist, including:

- `REVIEW:`-prefixed IRIs;
- unresolved `MISSING DESCRIPTION:` / `MISSING METADATA:` placeholders; and
- blank final `observation_unit_iri` values in `metadata/tables.csv`.

## 8) Run strict final validation

When the package is genuinely ready, switch to strict validation:

```{r}
validate_salmon_datapackage(pkg_path, require_iris = TRUE)
```

This should pass only when the package is actually publication-ready.

If strict validation still fails because a measurement term genuinely needs a
new shared or DFO-specific ontology term, that is not a bug in the workflow —
it means the package is **not finalized yet**. Keep the term-request plan with
the package, resolve the ontology decision, then run strict validation again.

## 9) Publish or share the finished package

Once strict validation passes:

- share the **whole package folder** (or a zip of it), not just
  `datapackage.json`;
- keep `metadata/*.csv` with the data files; and
- keep `term-request-plan.csv` only as working governance support, not as part
  of the canonical final package unless you want that audit trail to travel with
  it.

In short: **reload -> validate -> detect gaps -> route requests -> rebuild EDH
if needed -> strict validate -> publish**.
