---
title: "After Excel Review: Finalize and Publish Your Package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{After Excel Review: Finalize and Publish Your Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE, purl = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Overview

This guide starts **after** you already ran `create_sdp()`, opened the package in
Excel (or another spreadsheet editor), and saved your reviewed metadata back to
CSV.

If you have not created the package yet, start with the
[5-Minute Quickstart](metasalmon.html).

If you want to assemble `dataset.csv`, `tables.csv`, and
`column_dictionary.csv` manually instead of continuing from a reviewed package,
use [Publishing Data Packages](data-dictionary-publication.html).

This is the post-review path:

1. reload the reviewed package from disk;
2. validate the reviewed metadata and data structure;
3. re-run semantic suggestions only for fields that are still unresolved;
4. detect likely ontology gaps;
5. decide whether each missing term belongs in the shared salmon-domain
   ontology or the DFO-specific bucket;
6. produce a concrete list of term requests to file;
7. rebuild EDH XML if needed; and
8. run strict final validation before publishing.

## 1) Reload the reviewed package

```{r, purl = FALSE}
library(metasalmon)

pkg_path <- "fraser-coho-2023-2024-sdp"
pkg <- read_salmon_datapackage(pkg_path)

names(pkg$resources)
pkg$tables
```

`read_salmon_datapackage()` reloads the package from the canonical
`metadata/*.csv` files, so you are checking the same metadata you just reviewed
in Excel.

## 2) Run a review-state validation pass

Start with the package-level validator in **non-strict** mode:

```{r, purl = FALSE}
review_check <- validate_salmon_datapackage(pkg_path, require_iris = FALSE)
```

This catches package/data mismatches now, before you worry about final ontology
coverage. Since 0.2.6 the same call also enforces any **declared
`primary_key`** (duplicate or missing key values are hard issues) and warns
about every unresolved `MISSING METADATA:` placeholder — so a package fresh
from Excel review will usually warn until the placeholders are filled in.

If you also want the semantic details separated out, inspect the reloaded
package directly:

```{r, purl = FALSE}
review_semantics <- validate_semantics(pkg$dictionary, require_iris = FALSE)

review_semantics$issues
review_semantics$missing_terms
```

Use this pass to answer three questions:

- **Did the Excel edits break the package structure?**
- **Which required metadata is still a placeholder?** (the warnings preview them, deduplicated by file and field — a long list is truncated, so treat it as a pointer, not an inventory)
- **Which measurement rows still lack a final semantic term?**

Do **not** switch to `require_iris = TRUE` yet unless you believe the package is
fully finalized. Strict validation is the last gate, not the first one.

## 3) Re-run semantic suggestions for only the unresolved pieces

The package-root `semantic_suggestions.csv` file is the original evidence trail
from `create_sdp()`. After review, it is usually more useful to re-run
`suggest_semantics()` against the **current** package state so you only inspect
what is still unresolved.

```{r, purl = FALSE}
reviewed_dict <- suggest_semantics(
  df = pkg$resources,
  dict = pkg$dictionary,
  codes = pkg$codes,
  table_meta = pkg$tables,
  dataset_meta = pkg$dataset
)

remaining_suggestions <- attr(reviewed_dict, "semantic_suggestions")

remaining_suggestions |>
  dplyr::select(
    target_scope,
    target_row_key,
    target_sdp_field,
    dictionary_role,
    search_query,
    label,
    source,
    score
  )
```

That usually produces a much shorter shortlist than the original package export,
because rows you already resolved in Excel are no longer the target.

## 4) Detect likely ontology gaps

Now detect the places where `metasalmon` still cannot find a shared SMN match,
but did find useful fallback candidates elsewhere:

```{r, purl = FALSE}
gaps <- detect_semantic_term_gaps(reviewed_dict)

gaps |>
  dplyr::select(
    target_scope,
    target_row_key,
    target_sdp_field,
    dictionary_role,
    search_query,
    top_non_smn_label,
    top_non_smn_source,
    gap_detection_basis,
    llm_decision,
    llm_new_term_label,
    llm_escalated_from,
    placement_recommendation,
    placement_confidence
  )
```

This is the key post-review gap table:

- rows here are still unresolved;
- `top_non_smn_*` columns show the best non-SMN evidence the package found;
- `gap_detection_basis` distinguishes candidate evidence, a final LLM term-gap
  decision, or both; and
- `placement_recommendation` gives a first-pass routing guess.

When `suggestions` is omitted, `detect_semantic_term_gaps()` reads both the
`semantic_suggestions` and `semantic_llm_assessments` dictionary attributes.
Final `request_new_term` decisions are therefore included automatically, even
when candidate rows contain an SMN match. An unresolved `reject_shortlist`
escalation remains traceable through `llm_escalated_from`.

If you pass `suggestions` explicitly, only candidate evidence and the embedded
`llm_*` columns in that table are used. The dictionary's assessment attribute is
intentionally ignored in that form.

If you only want a narrower slice, filter by role or scope:

```{r, purl = FALSE}
gaps |>
  dplyr::filter(dictionary_role %in% c("variable", "property", "entity"))
```

## 5) Decide shared salmon-domain vs DFO-specific routing

Use this plain-English rule:

- **Shared salmon-domain ontology (`smn`)**: use this when the term describes a
  reusable salmon science concept that another program, region, or organization
  could reasonably use too.
- **DFO Salmon Ontology (`gcdfo`)**: use this when the concept is clearly tied to DFO policy,
  operations, internal workflow, local identifiers, program-specific statuses,
  or other context that is not broadly reusable outside that setting.
- **Profile term (`profile`)**: use this for a local, program, or organization
  profile when neither shared SMN nor DFO-wide GCDFO governance is appropriate.

A good default test is:

- if the term would still make sense in a non-DFO salmon dataset, lean **SMN**;
- if the term only makes sense because of a DFO or local program process, lean
  **DFO-specific**.

### How routing is decided

`render_ontology_term_request()` accepts `smn`, `gcdfo`, `profile`, `uncertain`,
and `skip` row outcomes. Its precedence is:

1. an explicit `scope_overrides` value for the row;
2. a forced `scope`;
3. a recognized `llm_new_term_namespace`; and
4. the deterministic placement heuristic.

Treat `llm_new_term_namespace` as evidence, not authority. A human reviewer must
still decide whether a DFO-specific proposal belongs in GCDFO, a narrower
profile, or nowhere. Non-interactive uncertainty becomes `skip`; interactive
use asks for a routing decision.

## 6) Produce a concrete term-request list

First render a dry, non-interactive plan and inspect every route:

```{r, purl = FALSE}
requests <- render_ontology_term_request(
  gaps,
  scope = "auto",
  ask = FALSE,
  profile_name = "local-program"
)

request_plan <- requests |>
  dplyr::select(
    dataset_id,
    table_id,
    target_scope,
    target_row_key,
    target_sdp_field,
    dictionary_role,
    search_query,
    top_non_smn_label,
    top_non_smn_source,
    gap_detection_basis,
    llm_new_term_namespace,
    placement_recommendation,
    placement_confidence,
    request_scope,
    ontology_repo
  )

request_plan
readr::write_csv(request_plan, file.path(pkg_path, "term-request-plan.csv"))
```

That gives you a concrete list of which package row still needs help, the best
fallback evidence, and the proposed route. Review `skip` rows and any route
inferred from `llm_new_term_namespace`.

To apply reviewed row-by-row decisions, pass an override vector. Edit
`reviewed_scopes` so every value is one of `smn`, `gcdfo`, `profile`, or `skip`:

```{r, purl = FALSE}
reviewed_scopes <- request_plan$request_scope
# reviewed_scopes[2] <- "gcdfo"

requests <- render_ontology_term_request(
  gaps,
  scope = "auto",
  ask = FALSE,
  profile_name = "local-program",
  scope_overrides = reviewed_scopes
)

requests |>
  dplyr::select(
    request_scope,
    ontology_repo,
    request_title,
    target_row_key,
    search_query
  )
```

The override vector must have length one or one value per gap row. Rendering
uses the active SMN and GCDFO repository templates and never submits an issue.

### Preview issue drafts

Preview both SMN and GCDFO payloads before any submission:

```{r, purl = FALSE}
issue_preview <- requests |>
  dplyr::filter(request_scope %in% c("smn", "gcdfo")) |>
  submit_term_request_issues(dry_run = TRUE)

issue_preview
```

The helper only posts after an explicit call with `dry_run = FALSE`; it never
submits during detection or rendering. Keep `confirm = TRUE` for curator review.
Profile rows remain local governance artifacts unless you provide a repository
workflow for them.

## 7) Retain semantic mapping and decomposition evidence when available

Use SSSOM for reviewed relationships between whole vocabulary concepts. That
includes an explicit `sssom:NoTermFound` row when a released PSC compound
measurement has no defensible match in another versioned vocabulary. Do not put
property/entity/unit decomposition columns into SSSOM.

```{r, purl = FALSE}
write_sdp_sssom(
  pkg_path,
  mapping_sets = reviewed_sssom_paths
)
validate_sdp_sssom(pkg_path)
```

Store the ordered property, entity, constraint, statistical-modifier, and unit evidence in
the separate SDP measurement-decomposition artifact:

```{r, purl = FALSE}
write_sdp_measurement_decompositions(
  pkg_path,
  decompositions = reviewed_components
)
validate_sdp_measurement_decompositions(pkg_path)
```

metasalmon does not invent either record by default: both require reviewed
inputs. When their manifests are present, later KNB planning validates them and
includes them automatically inside the canonical SDP ZIP. This preserves the
evidence as part of the package without claiming native I-ADOPT conformance or
turning the component rows into concept mappings.

## 8) Finalize metadata and rebuild EDH XML if needed

Once the metadata is final and every surviving IRI is deliberate, regenerate EDH
XML if your publication path needs it:

```{r, purl = FALSE}
write_edh_xml_from_sdp(pkg_path)
```

This is the reviewed-package wrapper around `edh_build_hnap_xml()`. It refuses
to rebuild while obvious review-state markers still exist, including:

- `REVIEW:`-prefixed IRIs; and
- unresolved `MISSING DESCRIPTION:` / `MISSING METADATA:` placeholders in
  `description`, `observation_unit`, and `table_label`.

A *blank* `observation_unit_iri` is not one of those markers and does not block
a rebuild; only a `REVIEW:`-prefixed value does.

## 9) Run strict final validation

When the package is genuinely ready, switch to strict validation:

```{r, purl = FALSE}
validate_salmon_datapackage(pkg_path, require_iris = TRUE)
```

This should pass only when the package is actually publication-ready.

If strict validation still fails because a measurement term genuinely needs a
new shared or DFO-specific ontology term, that is not a bug in the workflow —
it means the package is **not finalized yet**. Keep the term-request plan with
the package, resolve the ontology decision, then run strict validation again.

## 10) Build reviewed EML and preview KNB publication when needed

KNB publication requires EML-specific facts that the canonical SDP does not
guess, including structured parties, rights, methods, measurement scales,
missing-value meanings, and access intent. Record those reviewed decisions in
`metadata/eml-mapping.yml`, then build and validate EML 2.2.0:

```{r, purl = FALSE}
# Copy once, then replace every example value and checksum with reviewed facts.
file.copy(
  system.file(
    "extdata",
    "eml-mapping-template.yml",
    package = "metasalmon"
  ),
  file.path(pkg_path, "metadata", "eml-mapping.yml")
)
```

```{r, purl = FALSE}
write_eml_from_sdp(pkg_path)
```

That standalone call creates valid reviewed EML for inspection. The later KNB
dry run deterministically rebuilds it once more so the final EML also describes
each named canonical SDP artifact as an `otherEntity`.

Preview the exact immutable DataONE objects, identifiers, checksums, access
decision, and OAI-ORE relationships without reading credentials or making a
network request:

```{r, purl = FALSE}
publish_sdp_to_knb(
  pkg_path,
  public = FALSE,
  dry_run = TRUE,
  representation = "expanded"
)
```

That dry run targets the **KNB Test Node** (`urn:node:mnTestKNB`), which is
what `knb_environment` defaults to for a dry run. The golden path develops a
package against the test node first and posts to production once it looks good
there, so rehearse in that order:

1. credential-free **test dry run** (the call above)
2. optional authorized **test deposit**, with a `dataone_test_token`
3. fresh **production dry run**, naming the environment explicitly
4. optional private **production deposit**
5. separate **KNB Publish** action for public release and DOI

Test objects are never promoted to production: a production deposit is planned
separately, with production credentials, and mints its own identifiers.

```{r, purl = FALSE}
publish_sdp_to_knb(
  pkg_path,
  public = FALSE,
  dry_run = TRUE,
  representation = "expanded",
  knb_environment = "production"
)
```

Switching the environment switches the DataONE network, member node,
coordinating node, resolver, and catalog endpoint together — they never vary
independently — along with which credential is read. Test artifacts are written
under `publication/test/`, so the reviewed production `metadata/eml.xml` and
`publication/knb-manifest.json` are left exactly as they were.

Live KNB publication is a separate explicit action. It requires a short-lived
DataONE JWT in the process-local option belonging to the environment
(`dataone_token` for production, `dataone_test_token` for the test node — they
are separate credentials), an ORCID-authenticated subject matching the EML
metadata provider, an explicitly named `knb_environment`, and an explicitly
supplied `confirm = TRUE`. A live call never infers its environment. The
confirmation means both that the exact pending manifest is approved and that
the caller has authority to redistribute the package.

Keep the JWT out of scripts, YAML, manifests, and shell history. Enter it into
the current R process only and clear it immediately after the live call:

```{r, purl = FALSE}
# `dataone_token` is the production credential; the test node reads
# `dataone_test_token` instead.
options(dataone_token = rstudioapi::askForPassword("Short-lived DataONE JWT"))
on.exit(options(dataone_token = NULL), add = TRUE)

publish_sdp_to_knb(
  pkg_path,
  public = FALSE,
  dry_run = FALSE,
  confirm = TRUE,
  representation = "expanded",
  knb_environment = "production"
)
```

`public = FALSE` is the recommended review path, but it is not a server-side
draft: a live call creates persistent production KNB objects. Completion
requires authenticated byte-for-byte and SystemMetadata readback for every
object, anonymous denial for both object bytes and SystemMetadata, a complete
authenticated catalog graph, and zero planned PIDs in the anonymous catalog.
The reviewed private plan also explicitly sets DataONE replication to disabled
with zero replicas, and live readback must match that KNB-only policy. Changing
to `public = TRUE` is a separate release decision: it requires public
redistribution authority and explicitly requests three DataONE preservation
replicas; metasalmon never infers that decision from a private deposit.

An expanded plan contains four kinds of object: the original data resources
named by `tables.csv`, validated canonical SDP artifacts with their exact
package-relative paths, the EML science-metadata record, and the OAI-ORE map.
The EML record describes raw files as data tables and named SDP artifacts as
supplementary entities; ORE `prov:atLocation` values preserve the hierarchy for
exact reconstruction. A closed inventory includes SSSOM, ordered measurement
decompositions, observation structures, and manifest-declared
reproducibility records without scanning arbitrary files. There is no methods
artifact to deposit: since sdp-0.3.0 a package carrying a
`metadata/methods.csv` is refused outright, with a pointer to
`migrate_sdp_methods()`. The KNB-specific
`metadata/eml-mapping.yml`, its authorization evidence and party details, and
all mutable `publication/` receipts remain local and are not deposited.

KNB DOI minting is a later opt-in release action, not part of
`publish_sdp_to_knb()`. KNB assigns the DOI to the selected science-metadata
version when its Publish action makes that package public; it does not assign a
separate DOI to each raw or supplementary object. Leaving a deposit private is
therefore the practical review workflow. Editing after deposit creates new
immutable versions rather than overwriting bytes.

For a corrected or otherwise revised private version, keep the same
`series_key`, add a new safe `publication.revision_key` to the sidecar, and bind
the new dry run to the verified manifest for the preceding version. Build the
revision in a fresh versioned SDP directory; never reuse or overwrite the prior
package directory:

```{yaml, purl = FALSE}
publication:
  public: false
  revision_key: corrected-expanded-sdp-2026-08-04
```

```{r, purl = FALSE}
publish_sdp_to_knb(
  revised_pkg_path,
  public = FALSE,
  manifest_path = file.path(
    revised_pkg_path,
    "publication",
    "knb-manifest.json"
  ),
  revision_manifest = previous_manifest_path,
  dry_run = TRUE,
  representation = "expanded",
  knb_environment = "production"
)
```

A revision continues the environment it started in; a manifest from one
environment can never be revised into the other.

Review that new manifest before repeating the live call with
`dry_run = FALSE, confirm = TRUE`. An interrupted live call is resumable: reuse
the byte-identical package, revision manifest, and output manifest. A completed
upload whose coordinating-network catalog check is delayed returns
`published_pending_catalog`; rerunning the same call performs fresh checks and
does not create duplicate objects.

If KNB's separate Publish action later mints a DOI, KNB creates another
science-metadata version. metasalmon does not yet import that KNB-created
version into a local predecessor manifest. Do not base a later automated
revision on the older pre-DOI manifest; use KNB's editor for that next change or
wait for an explicit manifest-import workflow.

Before any plan is written, metasalmon rejects referenced rows whose `source`
or `ontology` label identifies them as a `review-candidate`. This is a
deliberately offline fail-closed check; it does not resolve public IRIs or by
itself prove that a vocabulary release was governed. The canonical
transformation record must separately pin the approved vocabulary release and
verify its public term IRIs before the SDP is deposited.

## 11) Publish or share the finished package

Once strict validation passes:

- share the **whole package folder** (or a zip of it), not just
  `datapackage.json`;
- keep `metadata/*.csv` with the data files; and
- keep `term-request-plan.csv` only as working governance support, not as part
  of the canonical final package unless you want that audit trail to travel with
  it.

In short: **reload -> validate -> detect gaps -> route requests -> build the
required export -> strict validate -> dry-run -> publish**.
