---
title: "Setup and Credentials"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Setup and Credentials}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

Use this page once before the homepage quick example or the [5-Minute Quickstart](https://dfo-pacific-science.github.io/metasalmon/articles/metasalmon.html).

It covers two things:

1. GitHub credentials so `remotes::install_github()` works reliably.
2. Optional LLM provider setup if you want `llm_assess = TRUE` later.

## 1) GitHub setup for installs and updates

`metasalmon` is installed from GitHub. Public installs often work without a token, but setting up a GitHub Personal Access Token (PAT) up front avoids rate limits and awkward auth failures.

### Create a GitHub token

Create a PAT in GitHub:

- <https://github.com/settings/tokens>

For public package install/update, a read-only token is enough. If you also plan to read private GitHub repositories later, make sure the token can access those repositories too.

### Store it in `~/.Renviron`

```{r}
file.edit("~/.Renviron")
```

Add:

```{r}
GITHUB_PAT="paste token here"
```

Restart R after saving.

### Install `metasalmon`

```{r}
install.packages("remotes")
remotes::install_github("dfo-pacific-science/metasalmon")
```

### If you also need private GitHub CSV access later

The install token and the package's private-repo CSV helpers are related but not identical workflows.

When you later want to use `read_github_csv()` against a private repository, run:

```{r}
ms_setup_github(repo = "your-org/your-private-repo")
```

That verifies private GitHub access for the repository you actually intend to read.

## 2) Optional LLM provider setup

The base quick example and quickstart do **not** require an API key.

Only do this setup now if you want to run:

- `create_sdp(..., llm_assess = TRUE)`, or
- `suggest_semantics(..., llm_assess = TRUE)`

If you also plan to pass PDF or Excel files through `llm_context_files`, install the optional readers once. HTML, DOCX, `.R`, `.Rmd`, and `.qmd` inputs work without extra reader packages:

```{r}
install.packages(c("pdftools", "readxl"))
```

### DFO internal: `chapi`

If you are on the DFO internal network or VPN, open:

- <https://chapi-dev.intra.azure.cloud.dfo-mpo.gc.ca/>

Then:

1. click the user icon in the bottom left,
2. open **Settings**,
3. click **Show** next to **API Keys**,
4. copy the key value.

Store it in `~/.Renviron`:

```{r}
file.edit("~/.Renviron")
CHAPI_API_KEY="paste key here"
```

Optional overrides:

```{r}
CHAPI_MODEL="ollama2.mistral:7b"
CHAPI_BASE_URL="https://chapi-dev.intra.azure.cloud.dfo-mpo.gc.ca/api"
```

### External users: OpenRouter

```{r}
file.edit("~/.Renviron")
OPENROUTER_API_KEY="paste key here"
```

`llm_provider = "openrouter"` defaults to `openrouter/free`.

### External users: OpenAI

```{r}
file.edit("~/.Renviron")
OPENAI_API_KEY="paste key here"
```

Then choose an explicit OpenAI model when you call the LLM review path, for example:

```{r}
suggested <- suggest_semantics(
  df = your_data,
  dict = your_dict,
  llm_assess = TRUE,
  llm_provider = "openai",
  llm_model = "gpt-4.1-mini"
)
```

## 3) Next step

After this setup is done, go back to:

- [Homepage quick example](https://dfo-pacific-science.github.io/metasalmon/)
- [5-Minute Quickstart](https://dfo-pacific-science.github.io/metasalmon/articles/metasalmon.html)
- [LLM Review With Context Files](https://dfo-pacific-science.github.io/metasalmon/articles/llm-context-review.html)

If you are continuing from a reviewed package later, use:

- [After Excel Review](https://dfo-pacific-science.github.io/metasalmon/articles/post-review-package-publication.html)
